### 2013-07-23 patchy-selection-paper.tex
WorkingWiki replaced project description

% \newcommand{\var}{\mathop{\mbox{Var}}}
% \newcommand{\Exp}{\mathop{\mbox{Exp}}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}

@@

The covergent evolution of similar phenotypes in response to shared
selection pressures is a testiment to the power of selection to
repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level with
dispate taxa converging to the same phenotype through parallel genetic
changes in the same pathway, genes, 
or even by precisely the same genetic changes. Such precise
convergence points to the conservation of function of the genes
underlying adaptive traits, 
sometimes over deep time scales \citep{deephomologypapers}. Convergence on a molecular level also suggests that the path taken by adaptation can sometimes be relatively constained as few changes could have resulted in the adaptive phenotype, or that relatively few changes that can confer the advantagous phenotype of interest were sufficiently free of deleterious pleiotropic consequences to contribute to adaptation.

@@

Convergent adaptation can also occur within species, with individuals within a species 
adapted to the same environment in parallel through different large effect changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes REFs.
All else being equal the evolution of convergent phenotypes is more
likely with a 
higher mutational input, i.e.\ when the mutational target and population sizes are larger \citep{}. 
Standing variation for a trait can also increase the probability of convergence, 
as multiple pre-existing alleles can spread from low frequency, 
if the newly beneficial alleles were previously not too deleterious \citep{Orr,Hermission}. 
The geographic distribution of populations can also shape the probability of parallel mutation within a species.
A widespread species is more likely to adapt by multiple mutations in parallel if dispersal 
is geographically limited, as subpopulations will adapt via new mutations rather than waiting for migration \citep{RalphCoop}.

%When multiple rare changes sweep up from low frequency in parallel within a population this has been termed a soft sweep.\\

Intuitively, convergence is also more likely 
when geographically separated populations adapt to ecologically similar conditions. 
The probability that convergent adaptations arise independently
before adaptations spread between the populations by migration
will be enhanced if these adaptive alleles are maladapted in intervening environments,
since such adverse conditions can strongly spatially restrict the spread of locally adapted alleles \citep{slatkin}.

For example, a variety of plant species have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{turnerarabidopsis,mimulus};
the alleles conferring heavy metal tolerance are thought to be disfavored off of these patches. 
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration,
to particular substrates, e.g.\ dark rock outcrops or white sand dunes \citep{benson}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is primarily dark on a number of black lava flows separated by much lighter soil \citep{dice}.
Strong selection appears to favour such crypsis \citep{owls},
so perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible from the dark pigmentation adaptation on different outcrops \citep{nachman}. 
Similar situations have been demonstrated in other mouse systems \citep{beachmice,peromyscus}.

@@

%In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
%where a continuous population was subjected to a selection pressure
%that was constant across the entire environment.

@@

%%%
\subsection{Introduction to a patchy model}
\label{ss:patchyspace}

Consider a population spread uniformly across a landscape. Within this environment there are patches of
habitat that individuals are initially poorly adapted to, surrounded by large areas that the population is well adapted to.

Suppose it takes only a single mutational change to create an allele ($B$) that adapts an individual to the
poor habitat type. This mutational change occurs at a low rate of
$\mu$ per chromosome per generation. We assume that this initially rare new
allele has fitness $1+s_B$ relative to the unmutated type ($b$) in the ``poor'' habitat patches,
and assume that it has fitness $1-s_m$ when rare in the intervening areas, with
$s_m>0$. To simplify matters we assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.
(In particular, we require $s_m\neq 0$.)
\plr{NOTE: changed over to $s_m>0$.}

We are interested in the establishment of mutations in the ``poor'' patches by either
migration or mutation, and so we are only interested in whether the allele
can escape initial loss by drift when rare. Thus we do not have to
specify the the fitness of the homozygote, only that its
fitness is such the dynamics of the  allele when it is rare are
determined by the fitness of the heterozygote. 
More general dominance will only make small corrections,
with the exception of the recessive case, which we will omit.
Mostly, we follow the literature in treating the diploid model as essentially haploid.
% This allows us to treat a diploid model as being essentially haploid for our purposes. \gc{TRUE?}  \plr{would change e.g. freqs at migration-selection balance}

For the sake of simplicity we will assume a constant haploid population density $\rho$, 
across both types of habitat (we will return to discuss this equal
density point later). We further assume that the
variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$ (the dispersal variance).

\paragraph{Previous work} 
We will make use of a number previous results from the literature. 
\citet{slatkin1973geneflow}, working in one dimensional deterministic framework, 
showed that if the physical width of the patch is less than $2 \tan^{-1} (\sqrt{s_m/s_B})$, 
then there is not a stable equilibrium with both $b$ and $B$ present,
so that $B$ cannot establish within the patch due to migrational swamping \citep[see also][ for a review]{Lenormand}.
\citet{barton1987establishment}, again working in one dimension,
showed that this critical patch size also held in a stochastic framework \citep[see also the work of][]{Polk}. 
For patches above this critical size, \citet{barton1987establishment}
found that the probability of establishment of a new mutant that appears at distance $x$ from the patch
decays exponentially with distance from the patch, with the scale given by $\sigma/\sqrt{s_m}$.
% mutations appearing further than a few multiples of $\sigma/\sqrt{2s_m}$ away from such a patch 
% have vanishingly small probability of becoming established within the patch. 
\citet{barton1987establishment} also showed that mutations appearing within the patch have probability of establishment
less than the probability for a panimictic population 
\citep[which we denote $p_e$, and often approximate by $2 s_B / \xi^2$][]{haldane,fisher};
and that the probability approaches $p_e$ as the size of the patch increases.
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{Maruyama,cherry}. 
\plr{wait, what result? that fixation is hindered by geographic structure?}

%Maruyama: On the fixation probability of mutant genes in a subdivided population*
%should we cite Cherry http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1462507/

%To make things tractable, we assume that the distances between the patches are large relative to migration distance $\sigma$,
%and that the disadvantage $s_m$ ensures that on the relevant timescale,
%the allele is very unlikely to fix locally in the regions where it is at a disadvantage.

%equal deme sizes?  general migration rates?

%%%%%
\subsection{Time to the establishment of a locally adaptive allele
  due to mutational influx}
\label{ss:patchymutation}

Consider first a single, isolated poor habitat patch of area $A$ in which no $B$ allele has yet become established. 
As we are interested in patches where local adaptation can occur,
we will assume that our patch is larger than the cutoff for local establishment, 
i.e.\ wider than $2 \tan^{-1} (\sqrt{s_m/s_B})$.

Let $p(x)$ be the probability that a new mutant $B$ allele arising at location $x$ 
relative to the patch fixes within the poor habitat patch.
The function $p(x)$ can be found in one dimension in terms of Jacobi elliptic functions,
XXX as shown in the Appendix, but the expressions are not particularly helpful.
The total successful mutational influx per generation is then given by the integral of $p(x)$ over the entire species range $\int \rho \mu p(x) dx$,
and depends in a complicated way on the relationship of patch width and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If the width of the patch is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to assume that $p(x) = p_e$ within the patch, and $p(x) = 0$ otherwise,
for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_B A / \xi^2$. 
This holds for large patches, in which the number of mutations arising within the patch is larger 
than those arising sufficiently nearby;
we examine this more generally via simulation in Appendix XXX.
\plr{In Appendix: we simulated fixation prob in a patch with gradual transition.  We have convincing figures.}

The rate at which mutations arise and colonize a patch of area $A$ is therefore
\begin{align} \label{eqn:mutrate}
  \mutrate = \rho \mu \int p(x) dx  \approx 2 s_B \rho A \mu / \xi^2.
\end{align}
If this rate is low,  then the number of generations until a mutation arises that
will become locally established within the patch is exponentially distributed with rate $\mutrate$.  
Assuming that once a mutation becomes established it quickly reaches its equilibrium frequency across the patch, 
the expected time it takes for our patch to become colonized by the $B$ allele is approximately $1/\mutrate$.

%%%%
\subsection{Time to the establishment of a locally adaptive allele due to migrational influx}
\label{ss:patchymigration}
Now suppose that there are two patches, each of area $A$, of the poor
habitat type separated by distance $R$. 
We will consider the case where the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
for the purpose of comparison ignoring the possibility that a $B$ allele arises independently by mutation in the first patch.
As above, any copy of the new allele appears at location $x$ relative to the second, unadapted patch 
has probability approximately $p(x)$ of fixing locally in this patch,
whether the allele is a new mutation or is descended from the occupants of the first patch.

Due to migration--selection balance, 
there will be a small number of $B$ alleles descended from those of the first patch
present in the regions where they are disadvantageous;
the expected frequency at distance $r$ may again be expressed in terms of Jacobi elliptic functions \citep[Appendix XXX][]{barton},
although the equations are again not particularly helpful.
This expected frequency is the \emph{time-averaged} occupation frequency,
i.e.\ the average frequency of the allele at each location over a long time period,
below we study the fluctuations about this mean.
More helpful than the exact expressions,
we know that if $q(r)$ is the expected proportion present at migration--selection balance at distance $r$ from the first patch
in the absence of the second patch, 
then for large $r$ this proportion decays exponentially, 
\begin{align} \label{eqn:eqfreq}
  q(r) \approx C \exp( -\sqrt{2 s_m} r / \sigma),
\end{align}
where $C$ is a constant depending on the geographic shape of the populations and the selection coefficients. 
(Probably $C=1/2$ is close enough for applications; see Appendix XXX.)
\plr{same appendix with sims as above}
\gc{Should we also use this approx. to get the approx. rate of establishing mutations
arising outside of the patch, above. Feel like in the mutation rate
part we could pull $s_B$ out front of the rate as well.} 
\plr{huh?}

This allows us to get a good sense of the time scale of adaptation by migration between separated patches, as follows.
First imagine that the new allele is not advantageous in the second patch,
so that the expected number of alleles present in a small area $\Delta$ at distance $r$ from the already established patch
is $\Delta \rho q(r)$.
As these copies are disadvantageous, they form a subcritical branching process, which die out fairly soon.
Now add to this local selective advantage by stipulating that any copy of the allele in the new patch
has ``extra'' offspring that are marked in some way, on average $(1+s_B)-(1-s_m)=s_B+s_m$ additional offspring per generation.
If the probability of establishment \emph{in the new patch} is $p(x)$ as above,
then each of these new, marked copies appearing at location $x$ has probability $p(x)$ of establishing.
(This general relation is stated formally in Appendix XXX.)
Therefore, the total rate of establishment 
(i.e.\ the mean number of successfully establishing migrant lineages per generation)
is $\rho (s_B+s_m) \int q(|x|) p(x) dx$.
Using a similar approximation as in section \ref{ss:patchymutation},
we can approximate this, for a patch of area $A$ at distance $R$ from an already occupied patch, as
\begin{align}
  \migrate(R) = \rho A \frac{s_B}{\xi^2} (s_B+s_m) \exp\left( -\frac{ \sqrt{2 s_m} R}{\sigma} \right),
\end{align}
where we have set $C=1/2$ in equation \eqref{eqn:eqfreq}.
Using different methods, we arrive at the same functional form below in section \ref{ss:haplotype_length}.
(Note that if the size of the new patch is large relative to $\sqrt{s_m}/\sigma$,
then the integral of $q(x)$ over the patch will not be well approximated by $A \times q(R)$,
but the adjustment is easy to make.)

If this establishment rate is low, the time it takes
a mutation to arise that will become established 
will be approximately exponentially distributed with rate $\migrate$.
Assuming that the time it takes for a allele from the first patch to
become established in the second patch is short,
then  the expected waiting time until some migrant allele fixes in the
second patch will be $1/\migrate$.

%%%
\subsection{The probability of parallel adaptation between patches.} 
\label{ss:probparallel}

We now turn to the question of whether the separated populations adapt by parallel genetic changes 
or by the spread of a migrant allele between the patches. 
% These two scenerios are assumed to be distinguishable, 
% either because in the case of parallelism the mutations occur at different positions 
% or if the mutations occur at the same position they can be distinguished due to their haplotypic background. 
% Assuming that the geographically spread population initally contains no copies of the $B$ allele, 
% one or other of the patches of different habitat 
% will be the first to adapt via the establishment of a $B$ allele;
% so the question is then the relative timescale of new mutation and migration to the other patch.
As only a single mutation is required to adapt individuals to the
poor habitat patch, subsequent mutations that arise after an allele becomes established on the patch gain no selective benefit. 
Similarly, an allele introduced into a patch by migration will not be favored by selection to spread, 
if the patch has already been colonized. 
Therefore, mutations selectively exclude each other from patches, over short time scales, 
and will only slowly mix together over longer time scales by drift.

We will assume that once a $B$ allele is introduced (by migration or mutation), 
it becomes established in the poor habitat patch rapidly. 
Under this approximation, and our selective exclusion assumption, 
after one of the patches becomes colonized by a single $B$ allele, 
the other patch will become colonized by migration or mutation, but not by both. 
As such, the question of whether the
population adapts to the second poor habitat patch in parallel, 
simply comes down to whether the a new mutation or a migrant allele is the first to become established in the second patch.

Taking expressions \eqref{eqn:mutrate} and \eqref{eqn:migrate} above,
we see that if the patches each have sufficiently large area,
the rate of migration and mutation will be on the same order if 
$\mu \approx ((s_B+s_m)/2) \exp(-\sqrt{2 s_m} R / \sigma)$,
where $R$ is the distance between the patches.
(Note that in both cases, $A$ is the area of the not-yet-adapted patch.)

If we take $\mu = 10^{-8}$, and $s_B=s_m=s$, then this is satisfied if $R/\sigma \approx (18+\log(s))/\sqrt{s}$,
so if the selective pressure is fairly strong (say, $s=.05$),
then the distance between patches must be about 70 times the dispersal distance (i.e.\ $R/\sigma \approx 68$);
while if $s$ is much smaller, say $s = .001$, 
then the patches must be separated by many hundreds of times the dispersal distance (i.e.\ $R/\sigma \approx 368$).
The situation does not change much if $\mu$ is larger -- even if $\mu = 10^{-6}$, 
the required separation between patches is only reduced by $5/\sqrt{s_m}$.
If the separation between patches is much larger than this value -- certainly, by a factor of ten -- 
then patches are effectively isolated -- most adaptation will occur independent between patches.
On the other hand, if the separation is much smaller, then a single
mutation is likely to fix in both patches.

If we are willing to take these approximations at face value, 
and assuming both rates are small, 
the time till the first patch is colonized by the $B$ allele will be approximately exponentially distributed with rate $2 \mutrate$.
Following this, the time till the second patch is subsequently colonized 
(via either migration or new mutation) 
will be exponentially distributed with rate $\mutrate+\migrate$.
Both these times scale linearly with $\xi^2/(\rho A s_{B})$, so that 
the overall time to adaptation is robustly predicted to increase with offspring variance ($\xi^2$)
and decrease with patch population size and benefical seleciton coefficient.
Furthermore, the probability that the second adaptation is a new mutation,
i.e.\ the probability of parallel adapatation, is
\begin{equation}
  \frac{\mutrate}{\mutrate+\migrate} = \frac{2\mu/(s_B+s_m)}{2\mu/(s_B+s_m) + \exp\left(- \sqrt{2 s_m} R / \sigma \right) },
\end{equation}
so that the probability of parallel mutation should increase
approximately logistically with the distance between the patches at rate $\sqrt{s_m} /\sigma$.

%%%%
\subsection{Multiple patches}

We can extend this reasoning to multiple patches of unadapted habitat 
in which the $B$ allele can facilitate local adaptation. 
Under our assumptions of mutational exclusion and rapid establishment, 
each patch will adapt through establishment of a single copy of the $B$ allele, 
either by migration or mutation.
Since we assume that the times between successive establishments is the result of many nearly-independent attempts
with small probability of success,
the process of occupation is well approximated by a discrete-state Markov chain,
whose state is the labeling of patches according to the colonizing allele
(or none, for those that haven't yet adapted).

>>>>

% \newcommand{\var}{\mathop{\mbox{Var}}}
% \newcommand{\Exp}{\mathop{\mbox{Exp}}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\sn}{sn}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}

@@

The covergent evolution of similar phenotypes in response to shared
selection pressures is a testiment to the power of selection to
repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level with
dispate taxa converging to the same phenotype through parallel genetic
changes in the same pathway, genes, 
or even by precisely the same genetic changes \citep{zhen2012parallel}. 
Such precise convergence points to the conservation of function of the genes
underlying adaptive traits, 
sometimes over deep time scales \citep{deephomologypapers}. 
Convergence on a molecular level also suggests that the path 
accessible to adaptation is sometimes relatively constrained, 
as would be the case if only a few changes confer the advantagous phenotype 
and are sufficiently free of deleterious pleiotropic consequences.

@@

Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genotypic changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes REFs.
All else being equal, such evolution of convergent phenotypes is more likely with a 
higher mutational input, that is, when the mutational target and population sizes are larger \citep{}. 
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
and standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

Intuitively, convergence is also more likely 
when geographically separated populations adapt to ecologically similar conditions. 
The probability that convergent adaptations arise independently
before adaptations spread between the populations by migration
will be larger if these adaptive alleles are maladapted in intervening environments,
since such adverse conditions can strongly spatially restrict the spread of locally adapted alleles \citep{slatkin1973geneflow,}.

One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{turner2010serpentine,mimulus};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches. 
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes \citep{benson1933concealing}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is primarily dark on a number of black lava flows separated by much lighter soil \citep{dice1940ecologic}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible from the dark pigmentation adaptation on different outcrops \citep{nachman2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism}.

In this paper, we study this situation, namely,
when a set of alleles provide an adaptive benefit in geographically localized patches, 
separated by (inhabited) regions where the alleles are deleterious.
\plr{or neutral?}
The main questions are:
Under what conditions is it likely that each patch adapts in parallel, through new mutation,
and when is it likely that migration carries these alleles between patches?
How could we tell, after the fact, which has occurred?
We work in a model of continuous geography,
using a variety of known results and new methods.
In section XX we \dots QUICK OUTLINE

@@

@@

%%%
\subsection{Model of a patchy landscape}
\label{ss:patchyspace}

Consider a population spread across a landscaper, 
within which are patches of habitat to which individuals are (initially) poorly adapted,
surrounded by large areas to which the population is well adapted.
(When we refer to ``patches'' it is to these pieces of habitat.)
Suppose it takes only a single mutational change to create an allele ($B$) that adapts an individual to the poor habitat type, 
a change that occurs at a (low) rate of $\mu$ per chromosome per generation. 
We assume that this initially rare new
allele has fitness $1+s_p$ relative to the unmutated type ($b$) in these ``poor'' habitat patches,
and assume that it has fitness $1-s_m$ when rare in the intervening areas, with $s_p, s_m > 0$.
To simplify matters, we assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.
(The case where $s_m=0$ requires a different approach, which we do not treat here.)

We are interested in the establishment of mutations in the ``poor'' patches by either
migration or mutation, and so are mainly interested in whether the allele
can escape initial loss by drift when rare. 
Therefore, we do not have to specify the the fitness of the homozygote; 
only that the dynamics of the allele when rare 
are determined by the fitness of the heterozygote. 
More general dominance will only make small corrections,
with the exception of the recessive case, which we omit.
For the most part, 
we follow the literature in treating the diploid model as essentially haploid.

For the sake of simplicity, we assume a constant haploid population density $\rho$, 
across both types of habitat (we return to discuss this equal density point later). 
We further assume that the variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance).
We also aim to describe the scenario where it is sensible to think of large, discrete patches;
the case of a fine-scale hetergeneous environment is treated \citep{elsewhere}.

\paragraph{Previous work} 
We will make use of several previous results from the literature. 
\citet{slatkin1973geneflow}, working in a one dimensional, deterministic framework, 
showed that if the physical width of the patch is less than $2 \tan^{-1} (\sqrt{s_m/s_p})$, 
then there is no stable equilibrium with both $b$ and $B$ present,
so that migrational swamping prevents $B$ from establishing \citep[see also][ for a review]{lenormand2002limits}.
\citet{barton1987establishment}, using general theory of \citet{pollak1966survival}, showed (again in one dimension) 
that this critical patch size also holds in a stochastic framework.
\citet{barton1987establishment} also showed that
for patches above this critical size,
the probability of establishment of a new mutant that appears at distance $r$ from the patch
decays exponentially with distance from the patch, with the scale given by $\sigma/\sqrt{s_m}$,
and that mutations appearing within the patch have probability of establishment
less than the probability for a panimictic population,
but approaching this as the size of the patch increases.
(This latter panmictic probability of establishment we denote $p_e$,
and often approximate by $2 s_p / \xi^2$ \citep{haldane,fisher}.)
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{maruyama1970fixation,cherry2003diffusion}.

%%%%%
\subsection{Establishment of a locally adaptive allele due to mutational influx}
\label{ss:patchymutation}

Consider first a single, isolated poor habitat patch of area $A$ in which no $B$ allele has yet become established. 
We first compute the time scale on which new $B$ mutations appear and fix in such a patch.
As we are interested in patches where local adaptation can occur,
we will assume that our patch is larger than the cutoff for local establishment, 
i.e.\ wider than $2 \tan^{-1} (\sqrt{s_m/s_p})$ \citep{slatkin1973geneflow}.

Let $p(x)$ be the probability that a new mutant $B$ allele arising at location $x$ 
relative to the patch fixes within the poor habitat patch.
As shown in Appendix \ref{apx:elliptic_integrals},
the function $p(x)$ can be found (in one dimension) in terms of Jacobi elliptic functions,
but the precise expressions are not particularly helpful.
The total successful mutational influx per generation is then given by the integral of $p(x)$ over the entire species range,
% $\int \rho \mu p(x) dx$,
and hence depends in a complicated way on the relationship of patch width and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If the width of the patch is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to assume that $p(x) = p_e$ within the patch, and $p(x) = 0$ otherwise,
for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_p A / \xi^2$. 
This holds for large patches, in which the rate at which mutations arise within the patch is larger 
than the rate at which they appear nearby;
we examine this more generally via simulation in Appendix \ref{apx:establishment_sims}.

The rate at which mutations arise and colonize a patch of area $A$ is therefore
\begin{align} \label{eqn:mutrate}
  \mutrate = \rho \mu \int p(x) dx  \approx 2 s_p \rho A \mu / \xi^2.
\end{align}
If this rate is low, then the time (in generations) until a mutation arises that
will become locally established within the patch is exponentially distributed with mean $1/\mutrate$.  
Assuming that once a mutation becomes established it quickly reaches its equilibrium frequency across the patch, 
the time scale on which new patches become colonized by the $B$ allele from new mutation is therefore approximately $1/\mutrate$.

%%%%%%%
\subsection{Establishment of a locally adaptive allele due to migrational influx}
\label{ss:patchymigration}

Now suppose that there are two patches, each of area $A$ and separated by distance $r$. 
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(so, ignoring the possibility that a $B$ allele arises independently by mutation in the first patch).
To determine this, we dissect the process by which an allele transits between the patches,
in the process obtaining other useful information about the genealogy.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at distance $r$ from the patch by $q(r)$.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(r)$ is the solution,
which can then be solved to give $q(r)$ in terms of Jacobi elliptic functions (see Appendix \ref{apx:elliptic_integrals}).
Asymptotics of these implies that for large $r$ the mean occupation frequency decays exponentially, 
\begin{align} \label{eqn:eqfreq}
    q(r) \approx C \exp( -\sqrt{2 s_m} r / \sigma) \qquad \text{for large $r$},
\end{align}
where $C$ is a constant depending on the geographic shape of the populations and the selection coefficients. 
In Appendix \ref{apx:patchy_sims} we argue that $C=1/2$ is about right for most parameter values,
and also that these asymptotics are correct in both one and two dimensions
(the differential equation is only analytically solvable in one dimension).

This expected frequency $q(r)$ is the \emph{time-averaged} occupation frequency,
or in other words, the total number of $B$ alleles found at distance $r$ across $T$ generations, divided by $T$.
Therefore, if we look at a small area of size $\epsilon$, 
then $q(r) \epsilon$ is equal to the probability of finding any $B$ alleles there,
multiplied by the mean number of such, if there are any.
This points to an important fact: 
close to the original patch, the ``equilibrium frequency'' $q(r)$ describes well the density of the $B$ allele at most times,
but far away from the patch, there are rare excursions of a family of $B$ alleles, interspersed by long periods of absence.
To compute the rate of establishment by such migration,
we use properties of subcritical branching processes and Brownian motion,
combined with \eqref{eqn:eqfreq},
to obtain an approximation that we argue is valid if the distance $r$ is large relative to $\sigma/\sqrt{s_p}$.

The arguments depends on the following decomposition of the genealogy of $B$ alleles.
First, fix some intermediate distance $r_0$ from the established patch,
and classify all $B$ alleles present at any time outside $r_0$ into families
if they share a common ancestor living outside $r_0$.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated by a collection of independent spatial branching processes.
(This can be made formal in the limit of large population density.)
We can extend these families by 
including any individual outside the original patch (but perhaps inside $r_0$) 
that has a parent outside $r_0$,
so that these branching processes are rooted at $r_0$, and are killed upon reaching the original patch.
This decomposition is illustrated in figure \ref{fig:sim_excursions}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(r) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $r$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate of at which \emph{families of} migrant $B$ alleles establish in a new patch at distance $r$, 
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(r) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $r$ ) }
    \end{split}
\end{align}
All terms except the ``outflux of families'' are simple calculations with spatial branching processes;
and we can solve for this outflux using the established form for \eqref{eqn:gestalt_q}.
Here $\migrate$ reflects the mean number of establishing migrant families per unit time,
rather than individuals,
because family establishment events may include more than one individual,
so individual establishment events are clustered in time,
and using their overall density to obtain a typical waiting time until the next event
will underestimate that time.

%%%%
\subsubsection{The genealogy of migrant families}

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ the allele far from the patches where it is established.
Consider one such family.
As noted above, we can model this by assuming that each individual
migrates, then reproduces, independently of all other family members.
The mean squared migration distance is $\sigma^2$, takes the migrant in a random direction,
and if migration takes the individual into a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring
(with mean $1-s_m$ and variance $\xi^2$),
then assigning spatial locations to each individual in the tree,
and then pruning any branches that happen to wander into a patch.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
backmigrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

Since each individual has on average $(1-s_m)$ offspring,
a family has on average $(1-s_m)^t$ members in generation $t$.
Let $p_e(t)$ be the chance of extinction by time $t$, and let $K_t$ be the (random) family size conditioned on nonextinction,
so that $(1-s_m)^t = (1-p_e(t))\E[K_t]$.
It turns out that by assuming the fate of each offspring is independent (i.e.\ that this forms a branching process),
and that the distribution of offspring numbers is not too heavy-tailed \citep[see][for details]{jagers1975branching},
that $K_t$ has a limiting distribution: $K_t \dconv K$ 
and that $(1-s_m)^{t}/(1-p_e(t)) \to \E[K]$ as $t \to \infty$.
In other words, the chance of survival to time $t$ is a constant multiplied by $(1-s_m)^t$,
and conditional on survival, the family size is fixed.

We can understand this quite concretely, by a method described rigorously by \citet{geiger1999elementary}.
When we condition on survival until $t$, we condition on existence of at least one lineage from time $0$ to time $t$.
Once given this lineage, nothing distinguishes any of the other reproduction events --
each individual in this lineage must give birth to at least one offspring,
and all other reproduction events are unconditioned.
In other words, the genealogy of a family that survives until $t$
can be decomposed into a ``trunk'' that lasts until $t$
sprouting independent ``branches'' that may or may not survive until $t$.
This process is depicted in Figure~\ref{fig:branching_decomp}A.
Concretely, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$
(so $\P\{Z_t = 0\} = p_e(t)$),
then
\begin{align}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$,
and each $W_s$ is an independent copy of the offspring distribution, conditioned on $W_s \ge 1$.

This suggests that we add in the spatial motion by following only the motion of the ``trunk''
(i.e.\ the red line in figure \ref{fig:branching_decomp}),
whose motion is is approximately Brownian, with variance $\sigma^2$ per generation.
(Recall $\sigma$ is the dispersal distance.)
Suppose that the new patch is at distance $r$.
The probability density for the first time such a Brownian motion $B_s$ hits $r$ is
\citep[XXX]{feller}
\plr{work in $d=2$}
% note $\sigma B_t \deq B_{\sigma^2 t}$ so $\P\{ \sup_{s\le t} \sigma B_s \le r \} = \P\{ \sup_{s \le \sigma^2 t} B_s \le r \}
\begin{align} \label{eqn:hitting_dist}
  \P\{ B_t=r \;\mbox{and} \; B_u<r \;\mbox{for all}\; u<t\} =  \frac{r}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{r^2}{2t\sigma^2}\right) .
\end{align}
Let $\tau$ be this random time at which a given migrant family hits the new patch,
with the convention that $\tau =\infty$ if the family never gets here.
We approximate the chance that a given family arrives at the new patch at time $t$, i.e.\ that $\tau=t$,
by the product of equation~\eqref{eqn:hitting_dist} and the probability that the migrant family survives until time $t$.
For convenience, define $\alpha := - \log(1-s_m)$ to be the rate of decrease between patches (note that $\alpha \approx s_m$),
and $\mu_K = \E[K]$ to be the limiting mean family size conditioned on survival,
so that $1-p_e(t) \simeq (1/\mu_K) \exp(-\alpha t)$.
Then, the probability distribution of $\tau$ is given by
\begin{align}
    \P\{ \tau \in dt \} =  \frac{1}{\mu_K} \frac{r}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{r^2}{2t\sigma^2} - \alpha t \right) dt  .
\end{align}
